---
title: "Hacky Hours Exercise: Palmer Penguins Practice"
date: "04/30/2025"
format:
  html:
    number-sections: false
    code-fold: true
    code-summary: "Answer"
---

::: {.callout-note icon=false title="About the Data"}
The [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/){target="_blank"} dataset comes from the `palmerpenguins` R package. It was created as an educational alternative to the classic iris dataset and provides real-world data collected by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php){target="_blank"} from the Palmer Station, Antarctica 
Long Term Ecological Research Network (LTER) site. The dataset includes data for 344 penguins from three different species, collected across three islands in the Palmer Archipelago, Antarctica."
:::

### 0. Load libraries + data

::: {.callout-tip title="A few reminders:"}
- **Before you begin, ask yourself**: WHERE AM I?
- Let your code b r e a t h e, align code vertically, and use a consistent naming convention!
- Use `#` to create a comment in an R Script
- Use `----` to create new "headers" in an R Script after a comment
- Pipe Operator shortcut: `Ctrl` + `Shift` + M
:::

Install the R package `palmerpenguins`, if you haven't already:

```{r}
#| eval: false
#| code-fold: false

install.packages("palmerpenguins")
```

These are following libraries we'll use for these exercises:

```{r}
#| code-fold: false
#| message: false

library(palmerpenguins) # access penguins data
library(skimr) # summarize data
library(tidyr) # allows us to use the pipe (%>%)
library(dplyr) # cleaning and wrangling functions
```

Access the penguin data by referencing `penguins` from the `palmerpenguins` package in your code or function arguments.

Use `palmerpenguins::penguins` to access the dataset directly from the `palmerpenguins` package without loading the package with `library()`.

```{r}
#| eval: false
#| code-fold: false
# if package is loaded:
View(penguins)

# without loading package:
View(palmerpenguins::penguins)
```


### 1. Explore your data

Call the following functions **in the Console** to get familiar with and explore your data:

```{r}
#| eval: false
#| code-fold: false

View() # opens data as a spreadsheet-style data viewer in the Source Pane
skimr::skim() # alternative to `summary()` providing broad overview of dataframe
names() # prints out all the names of an object
head() # returns first 6 rows of df; useful for large data
str() # returns structure of an R object; another alt to `summary()`
glimpse() # alt to `str()`; may be more readable for you
unique() # use in combo w/ subset operator ($) to see unique values in a column
dim() # returns dimensions (# rows, # cols) of df or use `nrow()` or `ncol()`
anyNA() # check for NAs
colSums(is.na()) # counts number of NAs in each column
```

### 2. Subsetting 

::: {.callout-warning icon=false title="Exercise A"}
Create a new R object called `annual_penguins`and use `dplyr::select()` to create a subset containing the columns: `species`, `island`, `sex`, and `year`.
:::

```{r}
annual_penguins <- penguins %>% 
  select(species,
         island,
         sex,
         year)
```

::: {.callout-warning icon=false title="Exercise B"}
In the `annual_penguins` data frame, use `dplyr::filter()` to create a subset containing only female penguins. You come up with a new object name!
:::

```{r}
female_penguins <- annual_penguins %>% 
  filter(sex == "female")
```

::: {.callout-warning icon=false title="Exercise C"}
Starting from the data frame you created in Exercise B, use `filter()` to further narrow or subset the data to only Adelie penguins.
:::

```{r}
female_adelie_penguins <- annual_penguins %>% 
  filter(sex == "female" & species == "Adelie")

# this is also valid
female_adelie_penguins <- annual_penguins %>% 
  filter(sex == "female") %>% 
  filter(species == "Adelie")

# ~ Note: ~ #
# You can combine multiple conditions inside a single `filter()` using & (and), or you can chain multiple `filter()` calls using the pipe (%>%). 
# Both approaches produce the same result.
# Chaining filters can sometimes improve readability, especially when applying several filtering steps
```

### 3. Calculating

::: {.callout-warning icon=false title="Exercise A"}
Create a new column using `dplyr::mutate()` that is the ratio of bill length to bill depth. Hint: Ratio = bill_length_mm / bill_depth_mm
:::

```{r}
penguins_bill_ratio <- penguins %>% 
  mutate(bill_ratio = bill_length_mm / bill_depth_mm,
         .after = bill_depth_mm)

# ~ Note ~ #
# Recall we can use `.after` or `.before` to specify the new col position
```

::: {.callout-warning icon=false title="Exercise B"}
Calculate the mean body mass (in grams) for each penguin species. To do this, you will be using two new functions: `dplyr::group_by()` and `dplyr::summarize()`. 

Use the `?group_by` and `?summarise` Help Pages or search online for examples if you get stuck.

The result should be a summary table with one row per species and the average body mass for each.
:::

::: {.callout-note title="summarize() vs summarise()" collapse=true}
You might notice that `dplyr` also has the function `summarise()`. Both `summarise()` and `summarize()` work exactly the same and are aliases for each other. The only difference is American vs English spelling - whichever one you use, be consistent!
:::

```{r}
penguins_mean_body_mass <- penguins %>%
  group_by(species) %>%
  summarize(mean_body_mass_g = mean(body_mass_g, na.rm = TRUE))

# ~ Reflections ~ # 
# How do the species compare in terms of average body mass?
# What happens if you only run the `group_by()` line without summarizing?
# We use `na.rm = TRUE` to ignore missing values. What happens if we don’t include it?
```


::: {.callout-warning icon=false title="Exercise C"}
Calculate the average flipper length in millimeters for penguins on each island. Again, you will be using the functions: `dplyr::group_by()` and `dplyr::summarize()`. 

The result should be a summary table with one row per species and the average flipper length in millimeters for each.
:::

```{r}
mean_flipper_length_by_island <- penguins %>%
  group_by(island) %>%
  summarize(mean_flipper_length_mm = mean(flipper_length_mm, na.rm = TRUE))

# ~ Reflections ~ # 
# How does flipper length compare across islands?
# How would the result change if you grouped by species instead of island?
```

### 4. Enhancing (Transforming)

::: {.callout-warning icon=false title="Exercise A"}
Consider what you learned about the size of the penguins from Exercises B & C from the Calculating section.

Add a new column called `size` that categorizes each penguin as "small", "medium", or "large" based on its species. Use `mutate()` and `case_when()` from `dplyr`.
:::

```{r}
enhanced_penguins <- penguins %>%
  mutate(size = case_when(
    species == "Adelie" ~ "small",
    species == "Chinstrap" ~ "medium",
    species == "Gentoo" ~ "large",
    TRUE ~ NA_character_  # all other species get value of NA
  ))

# ~ Note ~ #
# `NA` is used for missing values
# `NA_character_` is used when working with character columns to ensure the type of missing value is consistent with the column's data type 
# With `NA`, the data type is assumed to be logical and could result in an error
```






